Lowestcase and Uppestcase letters: Advances in Derp Learning

### 2021-03-20 lowercase/paper/lowercase.tex
More work on paper, getting there...

%

% TODO: SDF vs SFD typo factory!

@@

% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}

% \usepackage{chessfs}
\usepackage{adjustbox}

@@

\interfootnotelinepenalty=0

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

@@

\setlength{\footnotesep}{2em}

% \newcommand\comment[1]{}
\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

@@

@@

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.95 \linewidth]{futurda}
\caption{ Type specimen for the generated font {\bf Futurda}. This is
  the classic font Futura, run through the final, improved
  vector-based model (Section~\ref{sec:complicated}) to make each
  letter lowercase. The letterforms \letterform{A}--\letterform{Z}
  (top) become quite readable lowercase versions. The extra-lowercase
  \letterform{a}--\letterform{z} are also almost legible, but are
  mostly just scaled-down and screwed up versions of the lowercase
  letterforms. Could definitely imagine this appear in the
  ``distressed fonts'' category of a 10001 Free TrueType Fonts CD-ROM
  in the 1990s, though.
} \label{fig:futurda}
\end{figure*}

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{nans}
\caption{
  Divergent training after only 29 rounds. We have \nan{} total error
  (is that good?). The example in column one is an inversion example
  generated by the \makeuppercase\ model, which is why it also looks
  like the Tunguska event, just of the opposite sign. The other two
  are regular inputs, whose predicted outputs are black holes. Start
  over!
} \label{fig:nans}
\end{figure}

\subsection{Fiddly bits} \label{sec:fiddly}
The training error over time appears in
Figure~\ref{fig:sdfmergederror}. It looks like the ones I have seen in
machine learning papers, although I don't like to read other people's
papers because it just seems like spoilers. There are several
noticeable events in the curve, which came from me fiddling with the
parameters or network as it ran. Here are some of the things I did:

\paragraph{Vacuuming and culling.} Sometimes a node will just be
dead (basically never activates) or an edge weight will be nearly
zero. In these cases an equivalent, tidier network can be made by
dropping the node or edge. Periodically I would perform these
processes, sometimes feeling particularly choppy and removing like
10\% of the parameters at a time. If these parameters are truly
useless with no hope of recovery, we simply get faster training
because there's less work to do.

\paragraph{Widening.} The opposite thing is to introduce new nodes.
Adding nodes to hidden layers is pretty easy. The thing that worked
best for me is to increase the size of the layer by 10--15\%, where
each new node has random incoming weights and bias 0. Then for each
node on the next layer, I add edges to some subset of these new nodes
(again generally 10\% of them) with weight 0. Since this weight is
zero, the network computes the same function, but has new gradients
to explore (in practice, it then experiences some shock after a
few training rounds, but then quickly fine-tunes this away). More
parameters means slower training, but also more potential to learn
interesting functions, or overfit!

\paragraph{Deepening.} It's also possible to add layers once the
network is trained. This can be done anywhere, but I liked doing it on
the output layer because this gets the most direct feedback from the
training examples, and so it updates quickly and changes there are
easy to understand. If you append the identity matrix (new layer is
the same size as the previous; each node has weight 1.0 to its
corresponding node and 0.0 elsewhere) then this network computes
the same function but has new gradients to explore. Adding a layer
did seem to help unlock a new training regime
(Figure~\ref{fig:sdfmergederror}); subjectively it also reduced
some weird artifacts in the SDFs that the model used to predict (makes
sense; this most natural thing for this layer to do is learn how to
predict a ``correction'' from the old prediction, for example
by smoothing/sharpening it). This seems to be borne out by the
weights, which are also fun to look at (Figure~\ref{fig:lastlayer}).

\paragraph{Generating features.} On the other side, randomly sampling
pixels from the input SDF does work, but I superstitiously believed
that it might be better to have more spatially meaningful features. I
wrote a program to generate a bunch of random simple features (one
line/blob with positive weights, one line/blob with negative weights).
It then chooses a set of them that are both {\em good} (maximum
standard deviation on a sample of training data) and {\em different
  from one another} (redundant or even partially redundant features
are less valuable). It was nice to satisfy my supersition, and the
dark wizard of superstitious fiddling with neural networks in the
hope that they do the thing was pleased as well. The features are
at least handsome (Figure~\ref{fig:makefeatures}).

I assume these are all standard things that neural people do, but they
do better and smarter versions of them because they are willing to
read other people's papers instead of trying to figure things out from
scratch all the time. But you gotta occupy yourself somehow while it
crunches for a month.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{sdfmergederror}
\caption{ The training error for the SDF models. The red curve is the
  \makeuppercase\ model, which generally has a higher error rate
  (perhaps simply because uppercase letters usually have more pixels
  set) and blue is \makelowercase. The first few rounds have error
  that's off the charts, well above 100. The most dramatic event is
  around round 200,000, where I reduced the weight decay factor to
  $0.999995$ (from $0.9995$). I guess you just need more nines to be
  more reliable. There are some other visible peaks, which occur when
  I do things like remove nodes with very low weights or which are
  almost never activated (Section~\ref{sec:fiddly}). These momentarily
  increase error but it is easily fine-tuned away (e.g. by learning
  new biases). The peak at around 1.4M rounds is when I added a new
  layer to the end of the model, which does seem to create a new
  training regime (clear downward slope now); but this also
  significantly increases the training cost per round. Even after
  2,000,000 rounds, the network is still apparently improving, but at
  a speed of about 1 pixel per several weeks. Eventually the extremely
  strict SIGBOVIK deadlines mean you just have to call it done.
} \label{fig:sdfmergederror}
\end{figure}

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though that part of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). Nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom row kind of represents biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

@@

Making the GPU code faster has really high value (could save weeks of
waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
know) I found a good trick was to generate different OpenCL code with
constants baked for each layer (for example their size and
\verb+indices_per_node+); this allows some of the inner loops to use
faster tricks for e.g.~multiplication by a compile-time constant
instead of depending on an argument or data. I have different routines
for sparse and dense layers. It might even make sense to recompile the
kernels for other parameters that change over the lifetime of
training, like the learning rate. The {\tt fma} instruction (so named
for the physical law $F=MA$) is a bit faster than
\verb|potential += w * v|, and I guess the compiler can't do this
itself because of IEEE horrors. But like, who cares? In my opinion you
should be able to put it in ``fast machine learning mode'' where it
readily makes precision errors, with a command-line option like
\verb+--fml+. With all the tweaking, the easiest win was to use the
{\tt restrict} keyword on arguments to tell the compiler that the
input cannot alias the output, for example; this presumably helps
it schedule instructions better.

@@

\subsection{Tracing}
Automatically tracing bitmaps into vector form is no doubt a solved
problem, but I chose not to look at spoilers. Since we actually have a
signed distance field, we can build a tracing routine directly off of
that. The approach I took consists of three steps. First, I generate a
bitmap of the SDF (at its native size) using the threshold. I separate
the image into a nested tree of connected components in this pixel
space; each component knows its single parent and whether it is
``land'' (inside the shape) or ``sea''. Characters like \letterform{e}
need internal cutouts, which are represented by a different winding
order (clockwise or counter-clockwise) for the contour. Some of the
tricky cases in computing this tree structure are given in f

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9 \linewidth]{connected}
  \caption{

  } \label{fig:connected}
  \end{figure}

% autotracing in here, I think?

% screenshot UI where I can draw letters and get live
% predictions? Just for the video perhaps? I like the "cool s" joke?

@@

lorem ipsum (Figure~\ref{fig:perfect})

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
\caption{
%  XXX Why doesn't this caption come out the way I'm expecting?
%  (in amsart style... can't have two figure*s on the same page?)
   Type specimen for the generated font {\bf Perfect Hallucination}.
  TODO write a description!
} \label{fig:perfect}
\end{figure*}

@@

dolor sit amet

@@

% autotracing

@@

% cool s

@@

>>>>

@@

% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}
\usepackage{chessboard}
% \usepackage{chessfs}
% \usepackage{adjustbox}

@@

% lets me explicitly set a. or 1. etc. as enum label
% \usepackage{enumitem}

@@

\interfootnotelinepenalty=0
\setlength{\footnotesep}{2em}

\newcommand\sfrac[2]{\!{}\,^{#1}\!/{}\!_{#2}}

% XXX the collection of crap above causes a first page
% with just "1" on it (probably conflicting packages). This
% magic just drops the first page, which is of course not
% the right solution, but...
\usepackage{atbegshi}% http://ctan.org/pkg/atbegshi
\AtBeginDocument{\AtBeginShipoutNext{\AtBeginShipoutDiscard}}

@@

% shift-reduce conflict

@@

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.95 \linewidth]{futurda}
\caption{ Type specimen for the generated font {\bf Futurda}. This is
  the classic font Futura, run through the final, improved
  vector-based model (Section~\ref{sec:complicated}) to make each
  letter lowercase. The letterforms \letterform{A}--\letterform{Z}
  (top) become quite readable lowercase versions. The extra-lowercase
  \letterform{a}--\letterform{z} are also almost legible, but are
  mostly just scaled-down and screwed up versions of the lowercase
  letterforms. Could definitely imagine this appearing in the
  ``distressed fonts'' category of a 10001 Free TrueType Fonts CD-ROM
  in the 1990s, though.
} \label{fig:futurda}
\end{figure*}

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{nans}
\caption{ Divergent training after only 29 rounds. We have \nan{}
  total error (hard to say if that's good or bad?). The example in
  column one is an inversion example generated by the
  \makeuppercase\ model, which is why it also looks like the Tunguska
  event, just of the opposite sign. The other two are regular inputs,
  whose predicted outputs are black holes. Start over!
} \label{fig:nans}
\end{figure}

\subsection{Fiddly bits} \label{sec:fiddly}
The training error over time appears in
Figure~\ref{fig:sdfmergederror}. It looks like the ones I have seen in
machine learning papers, although I don't like to read other people's
papers because it just seems like spoilers, and reading is the
opposite of writing! There are several noticeable events in the curve,
which came from me fiddling with the parameters or network as it ran.
Here are some of the things I did:

\paragraph{Vacuuming and culling.} Sometimes a node will just be
dead (basically never activates) or an edge weight will be nearly
zero. In these cases an equivalent, tidier network can be made by
dropping the node or edge. Periodically I would perform these
processes, sometimes feeling particularly choppy and removing like
10\% of the parameters at a time. If these parameters are truly
useless with no hope of recovery, we simply get faster training
because there's less work to do. Speed is exhilarating!

\paragraph{Widening.} The opposite thing is to introduce new nodes.
Adding nodes to hidden layers is pretty easy. The thing that worked
best for me is to increase the size of the layer by 10--15\%, where
each new node has random incoming weights and bias 0. Then for each
node on the next layer, I add edges to some subset of these new nodes
(again generally 10\% of them) with weight 0. Since this weight is
zero, the network computes the same function, but has new gradients
to explore (in practice, it then experiences some shock after a
few training rounds, but then quickly fine-tunes this away). More
parameters means slower training, but also more potential to learn
interesting functions, or overfit! Danger is exciting!

\paragraph{Deepening.} It's also possible to add layers once the
network is trained. This can be done anywhere, but I liked doing it on
the output layer because this gets the most direct feedback from the
training examples, and so it updates quickly and changes there are
easy to understand. If you append the identity matrix (new layer is
the same size as the previous; each node has weight 1.0 to its
corresponding node and 0.0 elsewhere) then this network computes
the same function but has new gradients to explore. Adding a layer
did seem to help unlock a new training regime
(Figure~\ref{fig:sdfmergederror}); subjectively it also reduced
some weird artifacts in the SDFs that the model used to predict (makes
sense; this most natural thing for this layer to do is learn how to
predict a ``correction'' from the old prediction, for example
by smoothing/sharpening it). This seems to be borne out by the
weights, which are also fun to look at (Figure~\ref{fig:lastlayer}).
All problems in computer science can be solved by an additional
layer of indirection!

\paragraph{Generating features.} On the other side, randomly sampling
pixels from the input SDF does work, but I superstitiously believed
that it might be better to have more spatially meaningful features. I
wrote a program to generate a bunch of random simple features (one
line/blob with positive weights, one line/blob with negative weights).
It then chooses a set of them that are both {\em good} (maximum
standard deviation on a sample of training data) and {\em different
  from one another} (redundant or even partially redundant features
are less valuable). It was nice to satisfy my supersition, and the
dark wizard of superstitious fiddling with neural networks in the
hope that they do the thing was pleased as well. The features are
at least handsome (Figure~\ref{fig:makefeatures}). Creativity is
enriching!

I presume these are all standard things that neural people do, but they
do better and smarter versions of them because they are willing to
read other people's papers instead of trying to figure things out from
scratch all the time. But you gotta occupy yourself somehow while it
crunches for a month.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{sdfmergederror}
\caption{ The training error for the SDF models. The red curve is the
  \makeuppercase\ model, which generally has a higher error rate
  (perhaps simply because uppercase letters usually have more pixels
  set) and blue is \makelowercase. The first few rounds have error
  that's off the charts, well above 100. The most dramatic event is
  around round 200,000, where I reduced the weight decay factor to
  $0.999995$ (from $0.9995$). I guess you just need more nines to be
  more reliable. There are some other visible peaks, which occur when
  I do things like remove nodes with very low weights or which are
  almost never activated (Section~\ref{sec:fiddly}). These momentarily
  increase error but it is easily fine-tuned away (e.g. by learning
  new biases). The peak at around 1.4M rounds is when I added a new
  layer to the end of the model, which does seem to create a new
  training regime (clear downward slope now); but this also
  significantly increases the training cost per round. Even after
  2,000,000 rounds, the network is still apparently improving, but at
  a speed of about 1 pixel loss per several weeks. Eventually the
  extremely strict SIGBOVIK deadlines mean you just have to call it done.
} \label{fig:sdfmergederror}
\end{figure}

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though the remainder of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). Nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom row kind of represents biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

@@

Making the GPU code faster has really high value (could save weeks of
waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
know) I found a good trick was to generate different OpenCL code with
constants baked for each layer (for example their size and
\verb+indices_per_node+); this allows the cmopiler to use faster
tricks in inner loops for e.g.~multiplication by a compile-time constant
instead of depending on an argument or data. I have different routines
for sparse and dense layers. It might even make sense to recompile the
kernels for other parameters that change over the lifetime of
training, like the learning rate. The {\tt fma} instruction (so named
for the physical law $F=MA$) is a bit faster than
\verb|potential += w * v|, and I guess the compiler can't do this
itself because of IEEE horrors. But like, who cares? In my opinion you
should be able to put it in ``fast machine learning mode'' where it
readily makes precision errors, with a command-line option like
\verb+--fml+. With all the tweaking, the easiest win was to use the
{\tt restrict} keyword on arguments to tell the compiler that the
input cannot alias the output, for example; this presumably helps
it schedule instructions better.

@@

\subsection{Tracing}
Automatically tracing bitmaps into vector form is no doubt a solved
problem, but I chose not to look at spoilers. Since we actually have a
signed distance field, we can build a tracing routine directly off of
that. The approach I took consists of three steps. First, I generate a
bitmap of the SDF (at its native size) using the threshold. I separate
the image into a nested tree of connected components in this pixel
space; each component knows its single parent and whether it is
``land'' (inside the shape) or ``sea''. Characters like \letterform{e}
need internal cutouts, which are represented by a different winding
order (clockwise or counter-clockwise) for the contour. Some of the
tricky cases in computing this tree structure are given in
Figure~\ref{fig:connected}. Once I have this tree structure, I trace
each pixel mass recursively (Figure~\ref{fig:trace}). I find a pixel
on the edge, and then walk around that edge clockwise (simple case
analysis on the three pixels ahead of me). As I walk the edge, I look
at the normal (orthographic as we are doing orthography) of the edge
and see where it reaches the edge value on the SDF; this point (a
float) is output into the contour. I'm guaranteed to get back to where
I started. I recurse by negating the SDF (outside becomes inside) and
bitmap (land becomes sea), and reverse the winding order of the result
of recursion.

This gives me a perfectly fine line-based outline of the SDF's shape.
Since I output points at every pixel, sometimes these points are
inefficient (i.e. a series of colinear points on a straight line),
and sometimes they reflect sharp corners that are not aesthetic. So
I then take a second pass at each contour, and try fitting B\'ezier
curves to sequences of points while the error remains low. Again I
did fitting with a black-box optimizer, which is nice. However,
the function being minimized also needs to be able to find the
closest point on a B\'ezier curve to another point, and although
this can also be done easily with the black-box optimizer, nesting
an optimizer invocation inside another one proved to be way too slow.
I found an old algorithm in a book I owned and was stymied by as a
child~\cite{glassner1990graphics}.

\begin{figure}[ht]
  \centering
\includegraphics[width=0.6 \linewidth]{connected}
\caption{
  Some tricky cases to think about when generating the nested
  connected components, as the first step of tracing SDFs. Area
  {\bf 0} is the outside of the entire letterform, but note that
  we should include the top-left corner even though it is not
  reachable without leaving the bounds of the image. Area {\bf 1}'s
  parent is 0; it has two holes within it, Areas {\bf 2} and {\bf 3}.
  At the top left, the four pixel chunks making up area {\bf 4} are
  not actually connected, but they separate the hole which is
  child are {\bf 5}. This hole must have one parent, so it means
  that all four pixel chunks are part of the same area {\bf 4}.
} \label{fig:connected}
\end{figure}

\begin{figure}[ht]
  \centering
\includegraphics[width=0.95 \linewidth]{trace}
\caption{ Tracing the SDF from Figure~\ref{fig:sdf} into vector
  format. Left image shows the nested connected components. Middle
  image is the initial straight-line trace, and the right image
  shows the simplified contours using quadratic B\'eziers.
} \label{fig:trace}
\end{figure}

@@

All right, but why stop there??

... the fonts in the previous section were created by modifying the
case of existing letterforms, with mixed success. We can also do this
to any letter-like shape. I built a UI for drawing letters and seeing
them uppercased and lowercased (and then re-lowercased and
re-uppercased) live, but it's impossible to demonstrate in paper
form. It's pretty much what you'd expect.

The UI also tells you how much your input resembles the various
existing letters \letterform{A}--\letterform{Z} and
\letterform{a}--\letterform{z} using the 26 bonus outputs that each
model predicts. For example, I learned that the ``Cool
S''~\cite{wikipediacools}:

\begin{center}
\includegraphics[width=0.1 \linewidth]{cools}
\end{center}

\noindent does not much resemble an S.
% XXX ideally, numbers matching a scene in the video?

This begs us to ask the question: What shapes {\em do} look like
letters? Since the models will tell us, I can just search over shapes
and ask them. The first thing I tried was to just generate random
shapes and optimize their parameters to produce the highest possible
prediction for the target letter, and the lowest possible prediction
for the rest. This preduced results that are fully bonkers
(Figure~\ref{fig:bonkersf}).

We can improve the results by searching for inputs with a ``perfect''
prediction (1.0) rather than making it as high as possible. These
results may not have been fully bonkers, but were at least downright
wacky. Since there appear to be a large variety of inputs that the
model judges as ``perfect'', the most appealing results from this
excursion involved scoring some additional properties of the
hallucinated inputs to discourage them from being so barmy. I
generated $8 \times 8$ bitmaps, plenty of pixels to make readable
letters on classic computers. Rather than allowing them to be
arbitrarily noisy, I also weakly optimized for (1) the number of
pixels set being close to half (2) minimal transitions between
on and off along each row and column. This produced shapes that
are basically letter-like, but weird (Figure~\ref{fig:perfect}).

\begin{figure}[tp]
\centering
  \includegraphics[width=0.9 \linewidth]{bonkersf}
\caption{ A randomly generated SDF (left) and its rasterization
  (right) which maximized the predicted score for \letterform{F}
  (1.893 out of a nominal 1.0) while 
  scoring all other letters low. Parts are recognizable as an
  \letterform{F}, but other parts are fully bonkers. This is actually
  one of the least weird ones.
} \label{fig:bonkersf}
\end{figure}

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
\caption{
   Type specimen for the generated font {\bf Perfect Hallucination}.
   Each letter is an 8x8 bitmap that looks as close to ``perfect''
   as possible to the model. Perfect here means that for \letterform{C},
   the \makelowercase\ model outputs as close to the vector
   $[0, 0, 1, 0, 0, \ldots, 0]$ (in its 26 letter predictors; the
   actual lowercasing is ignored) as the optimizer could find. (Of
   course I didn't search all $2^{64}$ inputs, but errors are
   on the order of one part per thousand). The models are completely
   successful at recognizing normal-looking letters as well, but
   it likes these even better.
} \label{fig:perfect}
\end{figure*}

@@

@@

\newcommand\deterministic{
  \includegraphics[height=0.8em]{deterministic}
}
\newcommand\asymmetric{
  \includegraphics[height=0.8em]{asymmetric}
}

One obvious thing to do with a program that takes an $8 \times 8$
bitmap and produces some kind of score for it is to use that
program to play chess~\cite{murphy2019blind}. Here I entered
26 such programs in the Elo World tournament~\cite{murphy2019eloworld},
which allows us to see how they perform against each other
and benchmark algorithms. (Badly.)

At each turn, the algorithm takes the board state that would
result from each legal move, and interprets it as an $8 \times 8$
bitmap. It renders that bitmap as an SDF and then runs the
\makelowercase\ model on it, and chooses the move that minimizes
the difference between its letter predictions and the
$[0, 0, \ldots 1, \ldots 0]$ vector corresponding to the letter
we are ``playing as.'' \deterministic \asymmetric

The players all perform fairly badly. Letters \letterform{E} and
\letterform{S} are the worst (agreed on \letterform{S} being the worst,
thank you very much!), even worse than the ``No, I insist!'' strategy
that tries to force its opponent to capture its pieces. Letters
\letterform{K} and \letterform{T} (eponymous! yeah!) perform the best,
but still worse than random. The numeric players like $\pi$ and $e$
are categorically better than the alphabetical ones, but this is not
surprising because chess is more of a mathematical game than a
linguistic one (Figure~\ref{fig:chess}).

\begin{figure}[tp]
  \centering
  \chessboard[setfen=B2k4/8/1b2p2n/1b2p1pr/6Pp/4B2P/4q3/4K3 w - - 10 60,showmover=false]
  \caption{
    The letter \letterform{P} (white) loses to the numeric constant
    $\pi$ (black) after 59 nonsensical moves. The \letterform{P}
    player tries to move pieces such that the board looks like a
    letter \letterform{P} (and no other) to the neural network.
    The $\pi$ player uses
    $3 - \pi$ to arithmetically decode the sequence of legal moves
    (sorted alphabetically by PGN). Neither player is concerned with
    chess, really, but the letter-based players are generally bad
    because they are more likely to get stuck in local minima once
    they are basically happy with the shape of the pieces.
    %  1. Na3 g5 2. b3 e6 3. f4 Nc6 4. e4 Be7 5. e5 a6 6. c3 f6 7. h3 Bb4 8. g4 h5 9. Nf3 h4 10. Nd4 Rb8 11. Nf5 Na5 12. Nc2 Rh7 13. Nce3 d6 14. Qf3 Rh5 15. Nh6 Bd7 16. Bc4 c6 17. Rf1 Qc7 18. fxg5 Bxc3 19. b4 Bxb4 20. a3 Bc8 21. axb4 Qe7 22. Kd1 c5 23. d3 Nxh6 24. Ke2 Ra8 25. Rd1 dxe5 26. Bb5 Bd7 27. Ke1 c4 28. Nd5 Qg7 29. Nc3 Nb3 30. Bc6 Nd4 31. Be3 Nc2 32. Kf1 cxd3 33. Rdc1 d2 34. Qf4 Qh7 35. Qd4 Na3 36. Rxa3 a5 37. Qe4 b6 38. bxa5 dxc1=B 39. Qd5 Kd8 40. Qc5 Rb8 41. Ke1 Qe7 42. Bf3 fxg5 43. Bb7 Bxa3 44. Ba6 Bxc5 45. Bb5 Qd6 46. Ba6 Ba4 47. axb6 Bc6 48. Bc8 Qe7 49. Ba6 Rb7 50. Bxb7 Bxb6 51. Na2 Bb5 52. Nc1 Qa3 53. Nd3 Kd7 54. Nb2 Qxb2 55. Bc8 Ke8 56. Ba6 Kd7 57. Bc8 Kd8 58. Bb7 Qg2 59. Ba8 Qe2
    } \label{fig:chess}
\end{figure}

% figure? drop it?
{
  \tiny
\begin{tabular}{lrrrr}
name & elo & win & loss & draw \\
no i insist & 605.47 & 0 & 20405 & 42595 \\
letter f & 605.60 & 1341 & 21653 & 40006 \\
letter y & 606.00 & 1347 & 21688 & 39965 \\
letter b & 607.39 & 1705 & 21870 & 39425 \\
letter p & 607.61 & 1790 & 21963 & 39247 \\
letter l & 608.75 & 1370 & 21306 & 40324 \\
letter j & 610.86 & 1737 & 21525 & 39738 \\
letter u & 611.41 & 1363 & 20947 & 40690 \\
letter c & 612.38 & 1264 & 20787 & 40949 \\
huddle & 612.60 & 1172 & 20494 & 41334 \\
letter n & 612.81 & 1395 & 20802 & 40803 \\
letter w & 613.64 & 1342 & 20723 & 40935 \\
letter g & 613.72 & 1339 & 20660 & 41001 \\
letter v & 614.18 & 1390 & 20606 & 41004 \\
letter h & 615.04 & 1328 & 20452 & 41220 \\
letter r & 615.22 & 1826 & 20932 & 40242 \\
letter x & 615.70 & 1414 & 20457 & 41129 \\
letter m & 616.31 & 1774 & 20809 & 40417 \\
letter q & 618.85 & 1378 & 19993 & 41629 \\
letter o & 619.08 & 1502 & 20078 & 41420 \\
letter z & 620.09 & 1745 & 20275 & 40980 \\
letter d & 620.10 & 1730 & 20369 & 40901 \\
letter a & 620.57 & 1819 & 20212 & 40969 \\
letter i & 622.50 & 2169 & 20424 & 40407 \\
letter k & 622.90 & 1755 & 19878 & 41367 \\
letter t & 623.30 & 2237 & 20271 & 40492 \\
\ldots \\
random move & 655.90 & 7267 & 20983 & 34750 \\
\ldots \\
chessmaster.nes lv1 & 1014.74 & 37302 & 13992 & 11706 \\
\ldots
stockfish1m & 2831.66 & 60167 & 493 & 2340 \\
\end{tabular}
}

@@

@@

\begin{figure}[t]
\centering
  \includegraphics[width=0.7 \linewidth]{uppestcase3d}
\caption{ 3D manifold showing a section of the repeating loop as the
  \makeuppercase\ model iteratively uppercases a letter. (Shown here
  is the input \lowercase{q} from iteration 245--377, but they all
  converge to this same periodic shape.) Slices through this shape give a
  letterform's outline (or usually, a linear interpolation between two
  of them). The bottom of the shape is its ``beginning'' but it
  appears to repeat like this forever.
} \label{fig:uppestcase3d}
\end{figure}

% bristol stool chart
